---
title: "HDF"
output: 
  html_document:
    fig_caption: no
    number_sections: yes
    toc: yes
    toc_float: False
    collapsed: no
---

```{r set-options, echo=FALSE}
options(width = 105)
knitr::opts_chunk$set(dev='png', dpi=300, cache=FALSE, out.width = "75%", out.height = "75%")
pdf.options(useDingbats = TRUE)
klippy::klippy(position = c('top', 'right'))
```

# Introduction #

HDF (Hierarchical Data Format), see [[https://www.hdfgroup.org]](https://www.hdfgroup.org) is another format for storing large data files and associated metadata, that is not unrealted to netCDF, because in fact netCDF-4 uses the HDF-5 "data model" to store data.  There are currently two versions of HDF in use HDF4 and HDF5, and two special cases designed for handling satellite remote-sensing data (HDF-EOS and HDF-EOS5).  Unlike netCDF-4, which is backward compatible with netCDF-3, the two versions are not mutually readible, but there are tools for converting between the different versions (as well as for converting HDF files to netCDF files).  

HDF5 files can be read and written in R using the `rhdf5` package, which is part of the Bioconductor collection of packages.  HDF4 files can also be handled via the `rgdal` package, but the process is more cumbersome.  Consequently, the current standard approach for analyzing HDF4 data in R is to first convert it to HDF5 or netCDF, and proceed from there using `rhdf5`.

Compared to netCDF files, HDF files can be messy:  they can contain images as well as data, and HDF5 files in particular may contain groups of datasets, which emulated in some ways the folder/directory stucture on a local machine.  This will become evident below.

# Preliminary set up #

The `rhdf5` package can be retrived from Bioconductor, and installed as follows.  Generally, a lot of dependent packages will also need to be installed, as well, but the process is usually straightforward.

```{r install_rhdf5, eval=FALSE, echo=TRUE, message=FALSE}
# install rhdf5
if (!requireNamespace("rhdf5", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("rhdf5", version = "3.8")
```

The examples below use data from TEMIS, the Tropospheric Emission Monitoring Internet Service ([[http://www.temis.nl/index.php]](http://www.temis.nl/index.php)), and the Global Fire Emissions Database (GFED, [[https://www.globalfiredata.org/index.html]](https://www.globalfiredata.org/index.html)).  The TEMIS data are HDF4 and need to be converted, while the GFED data are HDF5.  The specific data sets can be found on the ClimateLab SFTP server.

The various conversion tasks and tools are:

- HDF4 to HDF5:  `h4toh5convert` from the HDFGroup [[h4h5tools]](https://portal.hdfgroup.org/display/support/Download+h4h5tools)
- HDF4 to netCDF:  `h4tonccf` (netCDF-3), and `h4tonccf_nc4` (netCDF-4), from HDF-EOS [[http://hdfeos.org/software/h4cflib.php]](http://hdfeos.org/software/h4cflib.php)
- HDF4 and HDF5 (and other file types) to netCDF:  `ncl_convert2nc` from Unidata [[https://www.ncl.ucar.edu/]](https://www.ncl.ucar.edu/overview.shtml)

The first two conversion tools have installers for MacOS, Windows and Linux.  On the Mac, it will usually be the case that the executable file should be moved to the /Applications folder, while on Windows the usual fooling around with paths will be required.  The `ncl_convert2nc` tool is part of NCL, which can be installed on run on Windows 10 machines using the Linux subsystem for Windows.

There are some useful tutorials on using HDF in R from the NSF NEON program:

- [[intro-hdf5-R]](http://neonscience.github.io/neon-data-institute-2016//R/intro-hdf5-R/)
- [[hdf5-intro-r]](https://www.neonscience.org/hdf5-intro-r)
- [[HDF5]](https://github.com/NEONScience/NEON-Data-Skills/tree/master/code/HDF5)

Begin by loading the appropriate packages:

```{r libraries, message=FALSE}
# load packages
library(sf)
library(rhdf5)
library(raster)
library(rasterVis)
```

For latter plotting of the data, also read a world outline shape file:
 ```{r shapefile}
# read a world shapefile
shp_path <- "/Users/bartlein/Projects/ESSD/data/shp_files/ne_110m_admin_0_countries/"
shp_name <- "ne_110m_admin_0_countries.shp"
shp_file <- paste(shp_path, shp_name, sep="")
world_shp <- read_sf(shp_file)
world_outline <- as(st_geometry(world_shp), Class="Spatial")

# plot the outline
plot(world_outline, col="blue", lwd=1)
```

## Convert an HDF4 file to HDF5

In this example, a data set of the "Clear-Sky UV Index"" from TEMIS was downloaded (i.e. `uvief20190423.hdf` -- 2019-04-23) from the website listed above.  This is an HDF4 file, which can be verified in Panoply.  Viewing files in Panoply is always good practice--in this case the particular layout of the data in file is rather ambiguous at first glance, and so getting a look at the contents of the file before attempting to read it is a good idea.

The HDF4 file can be converted by 

# Read a simple HDF5 file #

Begin by setting paths and filenames:

```{r, file1paths}
# set path and filename
hdf_path <- "/Users/bartlein/Dropbox/DataVis/working/geog490/data/hdf_files/"
hdf_name <- "uvief20190423.h5"
hdf_file <- paste(hdf_path, hdf_name, sep="")
```

List the contents of the file.  (WARNING:  This can produce a lot of output)

```{r listfile1}
# list the file content
h5ls(hdf_file)
```

Read the attributes of the `UVI_field` variable, the UV index:

```{r readattributes}
# read the attributes of UVI_field
h5readAttributes(hdf_file, name = "UVI_field")
```

Get the latitudes and longitudes of the grid:

```{r getlatlon1}
# get lons and lats
lon <- h5read(hdf_file, "/Longitudes")
nlon <- length(lon)
nlon
head(lon); tail(lon)

lat <- h5read(hdf_file, "/Latitudes")
nlat <- length(lat)
nlat
head(lat); tail(lat)
```

Get the UV data, and convert it to a `raster` object, and get a short summary:

```{r readh1}
# read the UV data
h1 <- h5read(hdf_file, "/UVI_field")
class(h1); str(h1)
UVI <- raster(h1)
UVI
```

Get a quick look using the `raster` version of the `image()` function:

```{r image1, fig.height=4, fig.width=4, out.width = "50%", out.height = "50%"}
# quick look
image(UVI)
```

The layout of the data in the file is obviously not straightforward.  First, the coordinates of the data are not longitudes and latitudes but simply values that range from 0 to 1, and the orientation of the grid is such that the `raster` object needs to be both tranposed (exchanging x and y), and flipped north to south.  this can be done as follows, also adding explicit longitude and latitude dimensions:

```{r transpose_and_flip}
# transpose and flip the raster 
UVI <- flip(raster(t(h1), xmn=lon[1], xmx=lon[nlon], ymn=lat[1], ymx=lat[nlat]), "y")
UVI
```

## Plot the UV data

A `rasterVis` plot can be produced as follows:

```{r rasterVis1}
# rasterVis plot
mapTheme <- rasterTheme(region=brewer.pal(8,"YlOrRd"))
plt <- levelplot(UVI, margin=F, par.settings=mapTheme, main="UV Index")
plt + layer(sp.lines(world_outline, col="black", lwd=1.0))
```

Close the open files before moving on:

```{r close1}
# close open HDF5 files
h5closeAll()
```

# Read a more complicated HDF5

A much more stuctured file that exploits the ability of HDF5 to store data in "groups" that mimics in some ways the directory or folder structure on a local computer is represented by the GFED4.1 fire data set.  As usual, begin by setting paths and file names.

```{r}
# set path 
hdf_path <- "/Users/bartlein/Dropbox/DataVis/working/geog490/data/hdf_files/"
hdf_name <- "GFED4.1s_2016.hdf5"
hdf_file <- paste(hdf_path, hdf_name, sep="")
```

List the file contents,  DOUBLE WARNING:  This really does produce a lot of output, only some is included here:

```{r list2, eval=FALSE, echo=TRUE}
# list the file contents
h5ls(hdf_file) # may create a lot of output
```

```{r output_listing, eval=FALSE, echo=TRUE}
##                            group                name       otype  dclass        dim
## 0                              /              ancill   H5I_GROUP                   
## 1                        /ancill       basis_regions H5I_DATASET INTEGER 1440 x 720
## 2                        /ancill      grid_cell_area H5I_DATASET   FLOAT 1440 x 720
## 3                              /           biosphere   H5I_GROUP                   
## 4                     /biosphere                  01   H5I_GROUP                   
## 5                  /biosphere/01                  BB H5I_DATASET   FLOAT 1440 x 720
## 6                  /biosphere/01                 NPP H5I_DATASET   FLOAT 1440 x 720
## 7                  /biosphere/01                  Rh H5I_DATASET   FLOAT 1440 x 720
## 8                     /biosphere                  02   H5I_GROUP                   
## 9                  /biosphere/02                  BB H5I_DATASET   FLOAT 1440 x 720
## 10                 /biosphere/02                 NPP H5I_DATASET   FLOAT 1440 x 720
## 11                 /biosphere/02                  Rh H5I_DATASET   FLOAT 1440 x 720
...

## 52                             /         burned_area   H5I_GROUP                   
## 53                  /burned_area                  01   H5I_GROUP                   
## 54               /burned_area/01     burned_fraction H5I_DATASET   FLOAT 1440 x 720
## 55               /burned_area/01              source H5I_DATASET INTEGER 1440 x 720
## 56                  /burned_area                  02   H5I_GROUP                   
## 57               /burned_area/02     burned_fraction H5I_DATASET   FLOAT 1440 x 720
## 58               /burned_area/02              source H5I_DATASET INTEGER 1440 x 720
## 59                  /burned_area                  03   H5I_GROUP                   
...
```

Read the longtidues and latiudes.  A preliminary look at the file in Panoply revealed that longitudes and latitudes were stored as full matices, as opposed to vectors.  Consequently, after reading, only the first row (for longitude) and first column (for latitude) were stored.

```{r readlatlon2}
# read lons and lats
lon_array <- h5read(hdf_file, "lon")
lon <- lon_array[,1]
nlon <- length(lon)
nlon
head(lon); tail(lon)

lat_array <- h5read(hdf_file, "lat")
lat <- lat_array[1,]
nlat <- length(lat)
nlat
lat <- lat[nlat:1]
head(lat); tail(lat)
```

Read the attributes of the `burned_area` variable:

```{r readattr2}
# read the attributes of the burned_area variable
h5readAttributes(hdf_file, name = "burned_area")
```

From the listing of the contents of the file, and the information provided in Panoply, it can be seen that, for example, the burned fraction (the proportion of a grid cell burned), for July 2016 is in the variable named `/burned_area/07/burned_fraction`.  Get that variable:

```{r readh2}
# read a raster slice
h2 <- h5read(hdf_file, "/burned_area/07/burned_fraction")
class(h2); str(h2)
BF <- raster(h2)
BF
```

Take a quick look:

```{r image2, fig.height=4, fig.width=4, out.width = "50%", out.height = "50%"}
# image plot
image(log10(BF))
```

The image plot, confirmed by Panoply, suggests that the raster needs to be transposed.

```{r transpose BF}
# tranpose the raster
BF <- raster(t(h2), xmn=lon[1], xmx=lon[nlon], ymn=lat[1], ymx=lat[nlat])
BF
```

## plot the GFED burned fraction data

Plot the data:

```{r rasterVis2}
# rasterVis plot
mapTheme <- rasterTheme(region=brewer.pal(8,"OrRd"))
plt <- levelplot(log10(BF), margin=F, par.settings=mapTheme, main="log10 Burned Fraction")
plt + layer(sp.lines(world_outline, col="black", lwd=1.0))
```


