---
title: "Long-term means and anomalies"
output: 
  html_document:
    fig_caption: no
    number_sections: yes
    toc: yes
    toc_float: false
    collapsed: no
---

```{r set-options_all, echo=FALSE}
options(width = 105)
knitr::opts_chunk$set(dev='png', dpi=300, cache=FALSE, out.width = "75%", out.height = "75%")
pdf.options(useDingbats = TRUE)
klippy::klippy(position = c('top', 'right'))
```

# Calculate long-term means and anomalies #

A common task in analyzing Earth-system science data is the calculation of "anomalies" or differences between individual months or years and some long-term average.  This produces two new data sets:  1) the long-term means ("ltm's") and 2) the anomalies ("anm's").  The example here gets the long-term means and anomlies for the CRU TS 4.02 near-surface air temperature data set `cru_ts4.02.1901.2017.tmp.dat.nc`.  The data include monthly values for the interval 1901 - 2017, and the long-term means will be calculated for one of the commonly used base periods, 1961 - 1990.

# Read the data #

Load the necessary packages, and set paths and filenames:

```{r}
# load the ncdf4 package
library(ncdf4)
# library(ncdf.tools)
library(lattice)
library(RColorBrewer)
#library(chron)

# set path and filename
ncpath <- "/Users/bartlein/Projects/ESSD/data/nc_files/"
ncname <- "cru_ts4.02.1901.2017.tmp.dat.nc"  
ncfname <- paste(ncpath, ncname, sep="")
dname <- "tmp"  # note: tmp means temperature (not temporary)
```

## Read dimensions and attributes of the data set ##

Open the netCDF file (and list its contents), and read longitudes, latitudes and times:

```{r}
# open a netCDF file
ncin <- nc_open(ncfname)
print(ncin)
```

```{r}
# get longitude and latitude
lon <- ncvar_get(ncin,"lon")
nlon <- dim(lon)
head(lon)
lat <- ncvar_get(ncin,"lat")
nlat <- dim(lat)
head(lat)
print(c(nlon,nlat))

# get time
time <- ncvar_get(ncin,"time")
tunits <- ncatt_get(ncin,"time","units")
nt <- dim(time)

nm <- 12
ny <- nt/nm

# convert time -- split the time units string into fields
print(tunits)
# tustr <- strsplit(tunits$value, " ")
# ptime <- convertDateNcdf2R(time, unlist(tustr)[1], origin = as.POSIXct(unlist(tustr)[3], 
#             tz = "UTC"), time.format = "%Y-%m-%d")
# head(time); tail(time)
# head(ptime); tail(ptime)
```

Get the years from `ptime` and also generate a sequence of index values for months.

```{r}
# get years from ptime & generate months
# year <- as.numeric(unlist(strsplit(as.character(ptime),"-"))[seq(1, nt*3, by=3)])
# head(year); tail(year)
# month <- rep(seq(1:12), ny)
# head(month); tail(month)
# head(ptime); tail(ptime)
```

## Read the array ##

Read the data, and get variable and global attributes:

```{r}
# get temperature
tmp_array <- ncvar_get(ncin,dname)
dlname <- ncatt_get(ncin,dname,"long_name")
dunits <- ncatt_get(ncin,dname,"units")
fillvalue <- ncatt_get(ncin,dname,"_FillValue")
dim(tmp_array)

# get global attributes
title <- ncatt_get(ncin,0,"title")
institution <- ncatt_get(ncin,0,"institution")
datasource <- ncatt_get(ncin,0,"source")
references <- ncatt_get(ncin,0,"references")
history <- ncatt_get(ncin,0,"history")
Conventions <- ncatt_get(ncin,0,"Conventions")
# head(ptime); tail(ptime)
```

Close the netCDF data set.

```{r}
# close the netCDF file
nc_close(ncin)
```

# Long-term means #

## Set up for long-term mean calculation ##

Replace netCDF fill values with R NA's

```{r}
# replace netCDF fill values with NA's
tmp_array[tmp_array==fillvalue$value] <- NA
length(na.omit(as.vector(tmp_array[,,1])))
```

Get a levelplot of a slice of the array to check.

```{r}
# levelplot of the slice
n <- 1
grid <- expand.grid(lon=lon, lat=lat)
cutpts <- c(-50,-40,-30,-20,-10,0,10,20,30,40,50)
levelplot(tmp_array[,, n] ~ lon * lat, data=grid, at=cutpts, cuts=11, pretty=T, 
  col.regions=(rev(brewer.pal(10,"RdBu"))))
```

Get the indices for the beginning and end of the base period, and save them as a string.
```{r}
# get beginning obs of base period
# begyr <- 1961; endyr <- 1990; nyrs <- endyr - begyr + 1
# begobs <- ((begyr - year[1]) * nm) + 1
# endobs <- ((endyr - year[1] + 1) * nm)
# base_period <- paste(as.character(begyr)," - ", as.character(endyr), sep="")
# print(c(begyr, endyr, begobs, endobs, base_period))
```

Get a levelplot of the first observation in the base period.

```{r}
# levelplot of begobs
# tmp_slice <- tmp_array[,, begobs]
# grid <- expand.grid(lon=lon, lat=lat)
# cutpts <- c(-50,-40,-30,-20,-10,0,10,20,30,40,50)
# levelplot(tmp_slice ~ lon * lat, data=grid, at=cutpts, cuts=11, pretty=T, 
#           col.regions=(rev(brewer.pal(10,"RdBu"))))
```

Create a new array with just the base period data in it.
```{r, eval=FALSE}
# base-period array
tmp_array_base <- array(dim = c(nlon, nlat, nyrs * nm))
dim(tmp_array_base)
tmp_array_base <- tmp_array[,, begobs:endobs]
```

Get a levelplot of that, which should match the previous plot

```{r, eval=FALSE}
# levelplot of tmp_array_base
tmp_slice <- tmp_array_base[,, 1]
grid <- expand.grid(lon=lon, lat=lat)
cutpts <- c(-50,-40,-30,-20,-10,0,10,20,30,40,50)
levelplot(tmp_slice ~ lon * lat, data=grid, at=cutpts, cuts=11, pretty=T, 
          col.regions=(rev(brewer.pal(10,"RdBu"))))
```

## Get long-term means ##

The long-term means are calculated by looping over the grid cells and months.

```{r, eval=FALSE}
# long-term means
tmp_ltm <- array(NA, dim = c(nlon, nlat, nm))
dim(tmp_ltm)
for (j in 1:nlon) {
  for (k in 1:nlat) {
    if (!is.na(tmp_array_base[j, k, 1])) {
      for (m in 1:nm)
        tmp_ltm[j, k, m] <- mean(tmp_array_base[j, k, seq(m, (m + nm*nyrs - 1), by=nm)])
    }
  }
}
```

Levelplot of the long-term means:

```{r, eval=FALSE}
# levelplot of tmp_ltm
tmp_slice <- tmp_ltm[,, 1]
grid <- expand.grid(lon=lon, lat=lat)
cutpts <- c(-50,-40,-30,-20,-10,0,10,20,30,40,50)
levelplot(tmp_slice ~ lon * lat, data=grid, at=cutpts, cuts=11, pretty=T, 
          col.regions=(rev(brewer.pal(10,"RdBu"))))
```


# Get anomalies #

The anomalies are gotten for each grid point by simply expanding the twelve ltm values (one for each month of the year) over the `ny` year of the record, and differencing.  In this way, the anomaly for each January in the record is the difference between the "absolute" (but not "absolute value" (`abs()`)) or observed value, and the January long-term mean.

```{r, eval=FALSE}
# anomalies
tmp_anm <- array(NA, dim = c(nlon, nlat, nt))
tmp_anm <- tmp_array - rep(tmp_ltm, ny)
```

Get a levelplot of an anomoly.  Note the different cutpoints.

```{r, eval=FALSE}
# levelplot of tmp_ltm
tmp_slice <- tmp_anm[,, 1]
grid <- expand.grid(lon=lon, lat=lat)
cutpts <- c(-10,-5,-2,-1,-0.5,0,0.5,1,2,5,10)
levelplot(tmp_slice ~ lon * lat, data=grid, at=cutpts, cuts=11, pretty=T, 
          col.regions=(rev(brewer.pal(10,"RdBu"))))
```

# Write netCDF files of the ltm's and anomalies ##

Write out netCDF files of the long-term means and anomalies in the usual way.

```{r, echo=TRUE, eval=FALSE}
# netCDF file of ltm's

# path and file name, set dname
ncpath <- "/Users/bartlein/Projects/ESSD/data/nc_files/"
ncname <- "cru_ts4.02.1961.1990.tmp.ltm.nc"   
ncfname <- paste(ncpath, ncname, sep="")
dname <- "tmp_ltm"  # note: tmp means temperature (not temporary)

# get time values for output
time_out <- time[(begobs + (nyrs/2)*nm):(begobs + (nyrs/2)*nm + nm - 1)]

# recode NA's to fill_values
tmp_ltm[is.na(tmp_ltm)] <- fillvalue$value

# create and write the netCDF file -- ncdf4 version
# define dimensions
londim <- ncdim_def("lon","degrees_east",as.double(lon)) 
latdim <- ncdim_def("lat","degrees_north",as.double(lat)) 
timedim <- ncdim_def("time",tunits$value,as.double(time_out))

# define variables
dlname <- "near-surface air temperature long-term mean"
tmp_def <- ncvar_def("tmp_ltm","degrees Celsius",list(londim,latdim,timedim),fillvalue$value,dlname,prec="single")

# create netCDF file and put arrays
ncout <- nc_create(ncfname,tmp_def,force_v4=TRUE)

# put variables
ncvar_put(ncout,tmp_def,tmp_ltm)

# put additional attributes into dimension and data variables
ncatt_put(ncout,"lon","axis","X") #,verbose=FALSE) #,definemode=FALSE)
ncatt_put(ncout,"lat","axis","Y")
ncatt_put(ncout,"time","axis","T")
ncatt_put(ncout,"tmp_ltm","base_period", base_period)

# add global attributes
ncatt_put(ncout,0,"title",title$value)
ncatt_put(ncout,0,"institution",institution$value)
ncatt_put(ncout,0,"source",datasource$value)
ncatt_put(ncout,0,"references",references$value)
history <- paste("P.J. Bartlein", date(), sep=", ")
ncatt_put(ncout,0,"history",history)
ncatt_put(ncout,0,"Conventions",Conventions$value)

# Get a summary of the created file:
ncout

# close the file, writing data to disk
nc_close(ncout)
```

```{r, echo=TRUE, eval=FALSE}

# netCDF file of anomalies

# path and file name, set dname
ncpath <- "/Users/bartlein/Projects/ESSD/data/nc_files/"
ncname <- "cru_ts4.02.1901.2017.tmp.anm.nc"   
ncfname <- paste(ncpath, ncname, sep="")
dname <- "tmp_anm"  # note: tmp means temperature (not temporary)

# recode NA's to fill_values
tmp_anm[is.na(tmp_anm)] <- fillvalue$value

# create and write the netCDF file -- ncdf4 version
# define dimensions
londim <- ncdim_def("lon","degrees_east",as.double(lon)) 
latdim <- ncdim_def("lat","degrees_north",as.double(lat)) 
timedim <- ncdim_def("time",tunits$value,as.double(time))

# define variables
dlname <- "near-surface air temperature anomalies"
tmp_def <- ncvar_def("tmp_anm","degrees Celsius",list(londim,latdim,timedim),fillvalue$value,dlname,prec="single")

# create netCDF file and put arrays
ncout <- nc_create(ncfname,tmp_def,force_v4=TRUE)

# put variables
ncvar_put(ncout,tmp_def,tmp_anm)

# put additional attributes into dimension and data variables
ncatt_put(ncout,"lon","axis","X") #,verbose=FALSE) #,definemode=FALSE)
ncatt_put(ncout,"lat","axis","Y")
ncatt_put(ncout,"time","axis","T")
ncatt_put(ncout,"tmp_anm","base_period", base_period)

# add global attributes
ncatt_put(ncout,0,"title",title$value)
ncatt_put(ncout,0,"institution",institution$value)
ncatt_put(ncout,0,"source",datasource$value)
ncatt_put(ncout,0,"references",references$value)
history <- paste("P.J. Bartlein", date(), sep=", ")
ncatt_put(ncout,0,"history",history)
ncatt_put(ncout,0,"Conventions",Conventions$value)

# Get a summary of the created file:
ncout

# close the file, writing data to disk
nc_close(ncout)
```
