---
title: "Plot Alpha"
output: html_document
---

# Introduction #

This RMarkdown document describes reading  a netCDF file consisting of several bioclimatic variables, and plots one of them `alpha`

## Load packages ## 

```{r load_packages, message=FALSE}
# load packages
library(sf)
library(raster)
library(rasterVis)
library(RColorBrewer)
```

# Read data #

## Shapefile ##

### Read a shapefile ###

Read a shapefile of world coastlines and countries.

```{r read_shp}
# read a world shapefile
shp_path <- "/Users/bartlein/Projects/ESSD/data/shp_files/ne_110m_admin_0_countries/"
shp_name <- "ne_110m_admin_0_countries.shp"
shp_file <- paste(shp_path, shp_name, sep="")
world_shp <- read_sf(shp_file)
world_outline <- as(st_geometry(world_shp), Class="Spatial")
```

### plot the shape file ###

```{r plot_shapefile}
# plot the world outlines
plot(world_outline, col="blue", lwd=1)
```

## Read alpha ##

Now read in alpha.

```{r read_data}
# read alpha (AE/PE)
alpha_path <- "/Users/bartlein/Projects/ESSD//data/nc_files/"
alpha_name <- "cru10min30_bio.nc"
alpha_file <- paste(alpha_path, alpha_name, sep="")
alpha <- raster(alpha_file, varname="mipt")
alpha
```

# Plot the data #

Finally, plot alpha using the `levelplot()` function from `rasterVis`:

```{r plot_data}
# rasterVis plot
mapTheme <- rasterTheme(region=brewer.pal(8,"BrBG"))
plt <- levelplot(alpha, margin=F, par.settings=mapTheme, main="AE/PE")
plt + layer(sp.lines(world_outline, col="black", lwd=1.0))
```