---
title: "Visualization for explanation"
output: 
  html_document:
    fig_caption: no
    number_sections: yes
    toc: yes
    toc_float: false
    collapsed: no
---

```{r set-options, echo=FALSE}
options(width = 105)
knitr::opts_chunk$set(dev='png', dpi=300, cache=TRUE, out.width = "75%", out.height = "75%")
pdf.options(useDingbats = TRUE)
klippy::klippy(position = c('top', 'right'))
```

# Introduction #

Although although visualization is often used simply to characterize individual variables, its use to understand relationship among variables is probably of greater value (see, for example, Wickham, H, and G. Grolemund, 2017, *R for Data Science*, O'Reilly [UO library](https://proquestcombo.safaribooksonline.com/9781491910382).  The examples here reproduce some of the analyses in Daniau, A.-L., P.J. Bartlein, S.P. Harrison and 57 others, 2012, Predictability of biomass burning in response to climate changes, *Global Biogeochemical Cycles* 26, GB4007, [doi:10.1029/2011GB004249](https://doi.org/10.1029/2011GB004249), which focued on both paleo and present-day relationship between global biomass burning and climate.  The present-day analysis focused on the prediction of satellite remote-sensing patterns of biomass burning from the GFEDv3.1 database (Global Fire Emissions, [http://www.globalfiredata.org](http://www.globalfiredata.org)).  The idea was to find a small number of predictors of the broadscale patterns of biomass burning, and it turned out that in both the paleo and present-day data sets, those predictors were mean annual temperature and precipitation minus evaporation (P-E, or `pme`).  The analysis below recapitulates some of the analyses in the Daniau et al. paper.

Here, most of the analysis is done using the `ggplot2` package for mapping and visualization.

# Data #

The input data consist of the GFEDv3.1 burned fraction (of half-degree grid cells, `gfed`), several bioclimatic variables (mean annual temperature (`mat`), mean temperature of the warmest and coldest months (`mtwa` and `mtco` respectively), growing degree days (5-deg C base, `gdd5`), annual precipitation (`map`), annual equilibrium evapotranspiration (`aet`), precipitation minus evapotranspiration (`pme`), lightning-strike frequency (`hmrc`), population density (`gpw`), net primary productivity (`npp`), treecover (`treecov`), and two categorical variables describing "megabiomes" and potential natural vegetation types (`megabiome` and `vegtype` respectively).

Load the necessary packages:

```{r}
# load packages
library(sf)
library(ggplot2)
library(tidyr)
```
 Read the data from a pre-made `.csv` file.
```{r}
# read the data
csvpath <- "/Users/bartlein/Projects/ESSD/data/csv_files/"
csvname <- "global_fire.csv"
csvfile <- paste(csvpath, csvname, sep="")
gf <- read.csv(csvfile)
str(gf)
summary(gf)
```

The two categorical (i.e. factor) variables have levels arranged in the default alphabetical order.  The following code reorders the levels into something more ecologically and climatically sensible:

```{r}
# reorder megabiomes
megabiome_name <- c("TropF", "WarmF", "SavDWd", "GrsShrb", "Dsrt", "TempF", "BorF", "Tund", "None", "Ice")
gf$megabiome <- factor(gf$megabiome, levels=megabiome_name)

# reorder vegtypes
vegtype_name <- c("TrEFW","TrDFW","TeBEFW","TeNEFW","TeDFW","BrEFW","BrDFW","EDMFW",
                  "Savan","GrStp","ShrbD","ShrbO","Tund","Dsrt","PDRI")
gf$vegtype <- factor(gf$vegtype, levels=vegtype_name)

# check the new ordering of factor levels
str(gf[16:17])
```

# Maps #

It's geneerally useful to map the data first.  The following code implements `ggplot2` maps.  A shapefile of world outlines is read the usual way with the `sf` package, then converted to a `ggplot2` compatible format using the `fortify()` function:

```{r}
# read a world outlines shape file
shp_path <- "/Users/bartlein/Projects/ESSD/data/shp_files/ne_110m_admin_0_countries/"
shp_name <- "ne_110m_admin_0_countries.shp"
shp_file <- paste(shp_path, shp_name, sep="")

# read the shapefile
world_shp <- read_sf(shp_file)
world_outline <- as(st_geometry(world_shp), Class="Spatial")
world_outline_gg <- fortify(world_outline)

# ggplot map of world_outline
ggplot(world_outline_gg, aes(long, lat)) + geom_polygon(aes(group = group), color = "gray50", fill = NA) +
  coord_quickmap() + theme_bw()
```

The map-pattern of the burned-fraction variable can be produced as follows:

```{r}
ggplot(gf, aes(lon, lat))  +
  geom_raster(aes(fill = log10(gf$gfed))) +
  scale_fill_gradient2(low = "white", mid="orange", high = "firebrick1", midpoint=-4, na.value="gray80") +
  geom_polygon(aes(long, lat, group = group), world_outline_gg, color = "gray70", fill = NA, lwd=0.5) +
  labs(title="GFEDv3.1 Burned Fraction") + 
  coord_equal() + theme_bw() + theme(legend.position="bottom")
```

# Set up #

## Transformation ## 

Simple inspection of the data and preliminary examination reveals that many of the variables have long-tailed distributions, and non-linear relationships.  This can be verified using a faceted plot of histograms.  Note the use of the `dplyr` `gather()` function to "stack" the variables, and the `free_x` argument to the `facet_wrap() function to allow each variable to plotted on individual as opposed to joint scales:

```{r, fig.width=6, fig.height=6}
# check distributions
ggplot(gather(gf[3:15]), aes(value)) + 
  geom_histogram(bins = 20) + 
  facet_wrap(~key, scales = 'free_x') +
  theme_bw()
```

For example, burned fraction has the kind of distribution, that if zeros are removed and the values tranformed by taking logarithms, would be approximately normal.

```{r}
# transform variables
hist(log10(gf$gfed[gf$gfed > 0]))
```

Find the minimum non-zero value of burned fraction:

```{r}
min(log10(gf$gfed[gf$gfed > 0]))
```

Tranform by taking logs (base 10), adding an increment slightly smaller than that minimum value, and check the distribution again.

```{r}
gf$gfed <- log10(gf$gfed + 1e-6)
hist(gf$gfed)
```

A few other variables can be transformed in a similar fashion:

```{r, eval=TRUE, echo=FALSE, results='hide'}
gf$hrmc <- sqrt(gf$hrmc)
min(log10(gf$gpw[gf$gpw > 0]))
gf$gpw <- log10(gf$gpw + 1e-10)
gf$map <- sqrt(gf$map)
gf$pme <- sqrt(gf$pme - min(gf$pme))
```
```{r, eval=FALSE, echo=TRUE}
hist(gf$hrmc)
gf$hrmc <- sqrt(gf$hrmc)
hist(gf$hrmc)

hist(log10(gf$gpw[gf$gpw > 0]))
min(log10(gf$gpw[gf$gpw > 0]))
gf$gpw <- log10(gf$gpw + 1e-10)
hist(gf$gpw)

hist(gf$map)
gf$map <- sqrt(gf$map)
hist(gf$map)

hist(gf$pme)
gf$pme <- sqrt(gf$pme - min(gf$pme))
hist(gf$pme)
```

Check the distributions again:

```{r}
# check distributions
ggplot(gather(gf[3:15]), aes(value)) + 
  geom_histogram(bins = 20) + 
  facet_wrap(~key, scales = 'free_x') +
  theme_bw()
```

## Censor observations with unusual megabiome values

In the following analyses, the `megabiome` variable will be used to label points and to construct faceted plots.  There are a few observations with `megabiome` values of "None" or "Ice":

```{r}
# check levels for megabiomes
levels(gf$megabiome)
table(gf$megabiome)
```

These can be removed by creating a list of the megabiomes we want to include, and dropping the "None" and "Ice" observations.  (The same could be done for the `vegtype` variable.)

```{r}
# drop last two categories
mb2 <- c("TropF", "WarmF", "SavDWd", "GrsShrb", "Dsrt", "TempF", "BorF", "Tund")
gf <- gf[gf$megabiome %in% mb2, ]
table(gf$megabiome)
```

# Controls of biomass burning #

There is a continuum of ideas on what controls biomass burning, which has at one end climate, and the other ignition (lighting and humans).  (It's envogue to always discuss biomass burning from the perspective of controls as a triangle, no matter what the scale of analysis, and so a triangle here would be climate-humans-lightning, but it turns out that humans and lightning aren't really significant at a global scale--lightning is well correlated with other climate variables, and humans suppress as many fires as they set.)

The goal here is to look at the relationships between burned fraction and the other available predictors.  The particular mechanism is a `ggplot2` plot, with added "smooths" (including both linear and non-linear (i.e. non-parametric) curves).  Although the addition of smooth curves to what are ordinary scatter plots could be viewed as simple plot-summary mechanisms, the are actually the first step in coming up with a systematic relationshp between one of the potential predictors and the response (burned fraction).  The following plots consist of simple scatter plots, supplemented by one of two smoothed curves:  1) a data-driven flexible curve, as implemented by a generalized additive model (GAM), which makes no assumption about the form of the relationship, other than it's relatively smooth; and 2) a linear (i.e. OLS) regression line.

## Bivariate relationship with burned fraction ##

```{r}
```

The first set of plots consist of one-predictor-variable-at-a-time plots, staring with the basic moisture and temperature variables as preditors.  Here, for example, is a plot of burned fraction vs. P-E, with a GAM curve added.  That curve is specified by the argument `(method = 'gam', formula = y ~ s(x, bs = "cs", k = 4` to the `geom_smooth()` function.  The `formula` argument specifies a GAM with a cubic spline (with "shrinkage"), and relatively small number of "knots" (`k=4`), which yields a relative smooth curve, one more appropriate form summarizing as opposed to interpolating).

```{r}
# single-variable plots
# burned fraction vs. pme
ggplot(data = gf) +
  aes(x = gf$pme, y=gfed) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs", k = 4), col="purple")  + 
  theme_bw()
```

The relationship is evidently nonlinear; burned fraction is highest at intermediate levels of P-E.  Here is a similar plot, with the Priestley-Taylor `alpha` (the ratio AE/PE) as a predictor:

```{r}
# vs. alpha
ggplot(data = gf) +
  aes(x = alpha, y=gfed) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs", k = 4), col="purple")  + 
  theme_bw()
```

The relationship is similar to that for P-E:  burned fraction is low in the driest *and* the wettest environments, and highest at intermediate levels of moisure.  

The other likely important predictor of biomass burning is mean annual temperature `mat`, which on a global scale controls biomass productivity, and hence fuel avaiabiliy. 

```{r}
# vs. mat
ggplot(data = gf) +
  aes(x = mat, y=gfed) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs", k = 3), col="purple") + 
  theme_bw()
```

The "curve" fit by the GAM model is not too curvy at all, and so it seems reasonable to summarize the relationship by a straight line  model:

```{r}
# vs. mat, linear smaoth
ggplot(data = gf) +
  aes(x = mat, y=gfed) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = 'lm', col="red")+ 
  theme_bw()
```


Other potential relationship can be explored as follows:

```{r, echo=TRUE, eval=FALSE}
# vs. gdd5
ggplot(data = gf) +
  aes(x = gdd5, y=gfed) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs", k = 3), col="purple") + 
  theme_bw()

# vs. treecov
ggplot(data = gf) +
  aes(x = treecov, y=gfed) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs", k = 3), col="purple") + 
  theme_bw()

# vs. gpw
ggplot(data = gf) +
  aes(x = gpw, y=gfed) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs", k = 3), col="purple") + 
  theme_bw()

# vs. hrmc
ggplot(data = gf) +
  aes(x = hrmc, y=gfed) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs", k = 3), col="purple") + 
  theme_bw()

# vs. npp
ggplot(data = gf) +
  aes(x = npp, y=gfed) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs", k = 4), col="purple") + 
  theme_bw() 
```

## Influence of vegetation type on biomass burning

Vegetation obviously has an impact on biomass burning, with low values in regions of sparse vegetation, like deserts and tundra, and potentially higher values in regions of dense vegetation.  The impact on the relationship between one of the potential predictors and burned fraction of vegetation type can be explored by point-labled plots or multi-panel plots.

```{r}
# points by vegtype
# burned fraction vs. alpha
ggplot(data = gf) +
  aes(x = alpha, y=gfed) +
  scale_color_brewer(palette="Accent") +
  geom_point(aes(color=gf$megabiome)) +
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs", k = 4), col="purple")  + 
  theme_bw()
```

```{r}
# multiplanel plots by vegtype
# vs. alpha
ggplot(data = gf) +
  aes(x = alpha, y=gfed) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs", k = 3), col="purple") + 
  facet_wrap(~ gf$megabiome, nrow=2) +
  theme_bw()
```
 (Note that in the above plot `alpha` is used ambiguously, both as a variable name and as a property of the plotted points (i.e. their tranparency).)
 
Other plots can be constructed:

```{r, echo=TRUE, eval=FALSE}
# vs. mat
ggplot(data = gf) +
  aes(x = mat, y=gfed) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs", k = 3), col="purple") + 
  facet_wrap(~ gf$megabiome, nrow=2) +
  theme_bw()

# vs. mat, linear smooth
ggplot(data = gf) +
  aes(x = mat, y=gfed) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = 'lm', formula = y ~ x, col="red") + 
  facet_wrap(~ gf$megabiome, nrow=2) +
  theme_bw()

# vs. npp
ggplot(data = gf) +
  aes(x = npp, y=gfed) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs", k = 3), col="purple") + 
  facet_wrap(~ gf$megabiome, nrow=2) +
  theme_bw()

# vs. npp, linear smooth
ggplot(data = gf) +
  aes(x = npp, y=gfed) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = 'lm', formula = y ~ x, col="red") + 
  facet_wrap(~ gf$megabiome, nrow=2) +
  theme_bw()
```

## A two-predictor Generalized Additive Model (GAM) ##

From the above, it seems that mean annual temperature `mat` and P-E `pme` (or AE/PE `alpha`) are the useful predictors of burned fraction (`gfed`).  That raises the possibility of including both in a predictive model, which can be implemented using the `mgcv` package (which was also implicity used above).

Load the `mgcv` library, and fit a model with burned fraction as the response, and `mat` and `pme` as predictors:

```{r}
# load the package
library(mgcv)

# regressions
gam01 <- gam(gf$gfed ~ te(mat, bs="cs", k=4) + te(pme, bs="cs", k=3), data=gf)
summary(gam01)
```

The first-order diagnostic plot for examining the fit of a GAM model can be gotten with the `gam.check()` function (not shown here)

```{r, echo=TRUE, eval=FALSE}
# diagnostic plots
par(mfcol=c(2,2))
gam.check(gam01)
par(mfcol=c(1,1))
```

An informative diagnostic is provided by the partial residual plots for each variable, which summarize the bivariate relationship between the response and the predictors.

```{r}
par(mfcol=c(1,2))
plot.gam(gam01, residuals=F, se=T, shade=T, pch=16, cex=0.5)
par(mfcol=c(1,1))
```

Finally, the bivariate relationship can be revealed by a 3-D response surface.

```{r}
# perspective plot
new_mat <- seq(from=-20, to=30, len=40); new_pme <- seq(from=0, to=80, len=40)
#new_mat <- seq(from=min(gf$mat), to=max(gf$mat), len=40); new_pme <- seq(from=min(gf$pme), to=max(gf$pme), len=40)
new_x <- expand.grid(mat=new_mat, pme=new_pme)
gam01_sfc <- predict(gam01, new_x)
gam01_sfc <- matrix(gam01_sfc, 40, 40)
persp(new_mat, new_pme, gam01_sfc, theta=-65, phi=30, d=1.5, col="gray", ticktype="detailed",
      zlim=c(-10,0), xlab="MAT (C)", ylab="sqrt(Annual P-E (mm))", xlim=c(-20,30), ylim=c(-10,80),
      zlab = "log10 GFED 3.1 Burned Fraction")
```

The take-home message here is that burned fraction responds linearly to mean annual temperature, and cuvilinearly with respect to P-E, with (as we have already seen) highest values at intermediate levels of moisture.