
---
title:  Raster mapping
output:
  html_document:
    css: html-md-01.css
    fig_caption: yes
    highlight: haddock
    number_sections: yes
    theme: spacelab
    toc: yes
    toc_float: false
    collapsed: no
---

```{r set-options, echo=FALSE, strip.white=TRUE}
options(width = 105)
knitr::opts_chunk$set(dev='png', dpi=300, cache=TRUE)
pdf.options(useDingbats = TRUE)
```

# Introduction #

This RMarkdown document describes reading a netCDF file consisting of several bioclimatic variables, and plots one of them, `alpha`.

## Load packages ## 

```{r alpha-plot-RMarkdown_Site-1 }
# load libraries
library(sf)
library(stars)
library(ggplot2)
```

# Read data #

## World outline ##

### Read map data from the `{maps}` package ###

Read a map file of  world coastlines and countries, and convert to an `sf` object.

```{r alpha-plot-RMarkdown_Site-2 }
# world_sf
world_sf <- st_as_sf(maps::map("world", plot = FALSE, fill = TRUE))
world_otl_sf <- st_geometry(world_sf)
```

### Plot the world outline ###

```{r alpha-plot-RMarkdown_Site-3 }
# ggplot map of world_outline
ggplot() + 
  geom_sf(data = world_otl_sf, fill = NA, col = "blue") + 
  scale_x_continuous(breaks = seq(-180, 180, 30)) +
  scale_y_continuous(breaks = seq(-90, 90, 30)) +
  coord_sf(xlim = c(-180, +180), ylim = c(-90, 90), expand = FALSE) +
  theme_bw()
```

## Read alpha ##

Now read in alpha from a netCDF file of bioclimatic variables using stars.

```{r alpha-plot-RMarkdown_Site-4 }
# read alpha (AE/PE)
# change path as necessary
nc_path <- "/Users/bartlein/Projects/RESS/data/nc_files/"
nc_name <- "cru10min30_bio.nc"
nc_file <- paste(nc_path, nc_name, sep="")
alpha <- read_ncdf(nc_file, var="alpha", proxy = FALSE)
```

List alpha to show contents of the object.

```{r alpha-plot-RMarkdown_Site-5 }
# list alpha
alpha
```

Convert the `stars` object to an `sf` object:

```{r alpha-plot-RMarkdown_Site-6 }
# convert stars to sf
alpha_sf <- st_as_sf(alpha, as_points = TRUE)
alpha_sf
plot(alpha_sf, pch = 16, cex = 0.3)
```

# Plot the data #

## Setup ##

Do some setting up befor plotting.

```{r alpha-plot-RMarkdown_Site-7 }
# setup for plotting alpha
# make a data.frame
lon <- st_coordinates(alpha_sf)[,1]
lat <- st_coordinates(alpha_sf)[,2]
alpha <- as.vector(alpha_sf)
alpha_df <- data.frame(lon, lat, alpha)
dim(alpha_df)
head(alpha_df)
```

Set axis labels or breaks:

```{r alpha-plot-RMarkdown_Site-8 }
# set axis labels (breaks)
breaks_x <- c(seq(-180, 180, by = 60)) 
breaks_y <- c(seq(-90, 90, by = 30)) 
labels_x <- as.character(breaks_x) 
labels_y <- as.character(breaks_y)
```

## Plot alpha ##

Make a `{ggplot2}` maps of alpha.

```{r alpha-plot-RMarkdown_Site-9 }
# ggplot2 map of alpha
ggplot() +
  geom_tile(data = alpha_df, aes(x = lon, y = lat, fill = alpha)) +
  scale_fill_gradient2(low = "brown", mid="white", high = "darkgreen", midpoint = 0.50) +
  geom_sf(data = world_otl_sf, col = "black", fill = NA) +
  coord_sf(xlim = c(-180, +175.0), ylim = c(-90, 90), expand = FALSE) +
  scale_x_continuous(breaks = breaks_x) +
  scale_y_continuous(breaks = breaks_y) +
  labs(x = "Longitude", y = "Latitude", title="Priestley-Taylor Coefficient (alpha) (AE/PE)", fill="alpha") +
  theme_bw()
```

# Discussion #

(A few paragraphs on map patterns, trends, etc.)